Explorations to Optimize the Parareal Algorithm \\to Solve ODEs in Given Applications

### 2014-12-10 paper.tex
expanding more on paper

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
We implement the parareal algorithm to solve various ODEs and compare in
terms of accuracy, speedup and efficiency. We explore the effect of using
different solution operators (a coarse and a fine as required by the method).
Our code is written in python using mpi4py to parallelize. We observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ and BLAS libraries which is a more
performance-oriented, lower-level language than Python. Data is shown for
solving for various complexities of ODEs from the basic exponential to
modeling sound wave propagation as we did in Homework 4.
\end{abstract}
%\end{@twocolumnfalse}
%]

@@

Another obstacle to trying to paralleliize numerical methods is that the methods
are inherently serial in time---in that evaluation of the next time $n+1$
depends on the previous values, say those at time $n-1$ and $n$. This setup
does not allow for parallelism. The only way to incorporate parallelism is to
create schemes that have parallelizable components---such as an update or a
refinement to a serial portion. As will be discussed and analyzed in the
metodology section, one such scheme would be the parareal algorithm.sFor an
overview, the parareal algorithm first does a coarse (low-order and fast)
approximation, then seeks to iteratively correct using smaller, more refined
numerical methods done in parallel from approximated starting points
approximated by the coarse solution.

An alternative paradigm to applying parallelism to numerical methods is the
parallelism by space paradigm, where the result vector space is divided up to
different regions and sent to different responsible processors. In this way,
each processor sees a similar but smaller-scale problem---weak scaling. This
method is expected to outperform the parallel in time paradigm simply because
the task is somewhat embarassingly parallel, being able to be naturally divided
into smaller but identical in method problems. There is no serial component in
this algorithm and should be exhibit both strong and weak scaling.

In general, the division in space paradigm is more intuitive to parallelize and
will often lead to greater speedups because there is no serial component. Yet
it is worth noting that knowing how to divide the result vector requires
knowledge of problem-specific details. Thus, this approach lacks the
general stability and theoretical derivations that schemes that parallelize in
time can offer.

@@

The parareal algorithm, developed by Lions, Maday and Turinici in 2001, CITE is
a general algorithm that allows parallelization in time.sIt does so by using a
cheaper (lower order, lesser resolution) approximation first, and then makes
corrections in parallel. The entire process is then iterative and tuned for a
given amount of iterations. Therefore, the algorithm is $k$ repetitions of a
serial coarse method updated by a finer method run in parallel for subsections.

@@

@@

@@

\subsection{Analysis of Speedup and Efficiency}

ashi TODO ashia from wesley's handwritten analysis

@@

We looked at both $e^x$ as well as TODO WHAT OTHER SCHEME using WHAT WERE THE PARAMETERS - Coarse and Fine Num Steps

@@

ashi TODO add figures and check my logic/bs to match it to the data

>>>>

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}

We look to achieve strong scaling in parallelizing numerical ODE methods. We test both parallelism by result space as well as parallelism by time, with the Parareal algorithm, to solve various ODEs.  We measure performance and compare in
terms of accuracy, speedup and efficiency. We provide some theoretical analysis for the Parareal algorithm and compare its tradeoff to paralleism by space, For the Parareal agorithm, we observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the Parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ using BLAS. Our code is written in python using mpi4py. We compare tests from various data sources from basic exponential ODEs to larger grids of ODEs to solve for in applications like diffusion and sound wave propagation.

\end{abstract}
%\end{@twocolumnfalse}
%]

@@

Another obstacle to trying to parallelize numerical methods is that the methods
are inherently serial in time---in that evaluation of the next time point $n+1$
depends on the previous values, say those at time $n-1$ and $n$. This setup
does not allow for parallelism. The only way to incorporate parallelism is to
create schemes that have parallelizable components---such as an update or a
refinement built on top of a more basic method first computed in serial. As will be discussed and analyzed in the
metodology section, one such scheme would be the parareal algorithm. As an
overview, the parareal algorithm first does a coarse (low-order and fast)
approximation, then seeks to iteratively correct using smaller, more refined
numerical methods done in parallel from approximated starting points
approximated by the coarse solution.

An alternative paradigm to applying parallelism to numerical methods is the
parallelism by space paradigm, where the result vector space is divided up to
different regions and sent to different responsible processors. In this way,
each processor sees a similar but smaller-scale problem---weak scaling. This
method is expected to outperform the parallel in time paradigm simply because
the task is embarassingly parallel, being able to be naturally divided
into smaller but identical in method problems. There is no serial component in
this algorithm and should be exhibit both strong and weak scaling.

In general, the division in space paradigm is more intuitive to parallelize and
will often lead to greater speedups because there is no serial component. The main tradeoff however, is that knowing how to divide the resultant system requires knowledge of problem-specific details. Thus, this approach lacks the
general stability and theoretical proofs (of error and convergence) that schemes which parallelize in
time can offer.

\subsection{Measuring Parallel Performance}

@@

The parareal algorithm, developed by Lions, Maday and Turinici in 2001, CITE is
a generalalized algorithm which allows for parallelization in time. It does so by using a
cheaper ($g_{\Delta t}$)---lower order or lesser resolution, approximation first which it then corrects by evaluating segments with a finer method ($g_{fine}$). The correction step from the approximation can be done in parallel. Multiple iterations can be performed to achieve an error provably asymptotically equal to a full serial computation of $g_{fine}$. The exact number of iterations required to achieve certain levels of convergence depends on the problem and must be tuned top optimize for speedup for a given system. In summary, the algorithm is $k$ repetitions of a
correction process which requires running updated by a finer method run in parallel for subsections.

@@

With the Parareal method, it is possible to combine ODE solvers. The stability region depends on both $g_{\Delta t}$ and $g_{\textnormal{fine}}$, and the
equation being solved.

@@

\subsection{Convergence of the Parareal}

TODO BKSIM

Assume the coarse operator $g_{\Delta t}$ is order $m$ and is Lipshitz, and the
fine solution operator $g_{\textnormal{fine}}$ is a sufficiently accurate
approximation to the analytic operator so we may replace $g_{\textnormal{fine}}
\to g$.

\emph{Theorem}: The order of accuracy of the Parareal method with coarse
solution operator $g_{\Delta t}$ and fine operator $g$ is $mk$. Can be proved by
induction.

SHALL WE DO IT THEN?

@@

\subsection{Parallel Tradeoff Analysis}

ashi TODO

Let's say the coarse operator runs in time $t$, and the fine operator runs in
time $Q \cdot t$. Assume we have $N$ processors and we perform $k$ correction
steps. Then, the runtime of Parareal is, assuming negligible setup and
aggregation time,

\begin{equation}
t + k(t + \frac{Qt}{N}).
\end{equation}

The first t seconds comes from the first coarse approximation, without which the parareal algorithm degenerates to $g_{\Delta t}$,  In order for there to be a speedup relative to the fine operator, we require
that $t + k(t + Qt/N) < Qt$, or $k < \frac{Q - 1}{1 + Q/N}$ or $N > \frac{Qk}{Q
- 1 - k}$. This can be tuned with by either adjusting the the quality factor or reducing the number of iterations, but the number of iterations, k, is important for error convergence to that of $g_{fine}$.  However, this optimal k is a problem-specific parameter and must be optimized for.

@@

We implemented the Parareal algorithm in Python, using mpi4py to parallelize it.
Our coarse operator was forward Euler step, with 100 steps, while the fine
operator was forward Euler, with $Q * 100$ steps, where $Q$ is the quality
factor. We tested the Parareal algorithm on two sets of differential equations:

\[
y'(t) = f(t, y) = \lambda y, \, y(0) = 1
\]
\[
y''(t) + 2y'(t) + 5y(t) = 0, \, y(0) = 1 \, y'(0) = 0.
\]

We ran the algorithm on different numbers of processors and varied $k$ and $Q$.

@@

TODO ashi sort out the figure and match and replace if needbe

\begin{figure}
\includegraphics[width=0.75\textwidth]{data/runtime_vs_cpus.pdf}
\caption{Running time vs. number of CPUs.}
\label{fig:run_v_cpu}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{data/error_vs_corrections.pdf}
\caption{Error vs. number of correction steps.}
\label{fig:err_v_k}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{data/error_vs_qualityfactor.pdf}
\caption{Error vs. quality factor of the fine operator}
\label{fig:err_v_q}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{data/runtime_vs_corrections.pdf}
\caption{Running time as a function of the number of correction steps.}
\label{fig:run_v_k}
\end{figure}
